RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2014-11-29 manuscripts/manuscript.Rmd
Edits for Reading Character Data.
Less extensive edits here, mostly to tighten text and improve flow.

Like NEXUS files, NeXML documents can in addition to trees also
contain character data in the form of matrices.  Character data can be
of molecular (molecular sequence alignments), discrete (most
morphological character data), or continuous type (many trait data). 
Often character data, especially molecular and morphological, is used
for phylogenetic tree reconstruction, but they are also a common input
for the many comparative phylogenetic analysis methods available in R. Methods to extract character data
work much like those for extracting the phylogenetic trees. Here we read in an example
NeXML file that contains both a phylogeny and character data. The
function
`get_characters()` can then be used to obtain a `data.frame` with columns as characters and rows as
taxa:

```{r }
nex <- nexml_read("components/comp_analysis.xml")
get_characters(nex)
```
Continuous and discrete character data are represented as separate
blocks in NeXML files, but will be combined as separate columns of a
single data frame by `get_characters()` if they correspond to the same
taxa. If this behavior is undesired, the function
`get_characters_list()` can be used to obtain
a list of data frames, where list elements correspond to the character matrix blocks of
the NeXML file.

>>>>

NeXML documents can contain character or trait data in the form of
matrices. A data matrix can be of molecular (for molecular sequence
alignments), discrete (for most morphological character data), or
continuous type (for many trait data). Character data, especially
molecular and morphological matrices, are often used for phylogenetic
tree reconstruction, but they are also a common input for the many
comparative phylogenetic analysis methods available in R. The method
`get_characters()` obtains character data from a `nexml` object, and
returns them as a standard `data.frame` R object with columns as
characters and rows as taxa. The following example, which uses a file
that comes with `RNeXML`, demonstrates this:

```{r }
nex <- nexml_read("components/comp_analysis.xml")
get_characters(nex)
```

To allow strict validation of data types, the NeXML schema requires
continuous and discrete character data to be separated into different
"characters" blocks. Since the `data.frame` data structure in R has no
such constraints, the `get_characters()` method will combine such
blocks as separate columns into a single `data.frame` object, provided
they correspond to the same taxa.
<!-- does the example data file actually demonstrate this? if not, it -->
<!-- should. HL -->
Otherwise, a list of `data.frame`s
is returned, with list elements corresponding to characters
blocks. Similar to the distinction between `get_trees()` and
`get_trees_list()`, there is also a method `get_characters_list()`,
which always returns a list of `data.frame`s.
